Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-08-01 review_article.Rmd
cofactors
editing introduction

Maps have been adopted as a base to present geospatial statistics for centuries.
These maps connect data and statistics to the geographical representation of areas that are generally familiar to the audience.
However, it is not enough for areas on maps to be recognisable, if information regarding the distribution of the statistic cannot be gleaned.
This is particularly relevant to cancer statistics, as cancer outcomes directly relate to the people living within a geographical area.
In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer statistics.
This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes, and health outcomes more broadly.

@@

<!--
Disease maps present geographic patterns that may have been overlooked in a table that obscures the geospatially related statistics [@SAMGIS].
Often used as a first step to drive hypotheses

Transforming statistics to visualisations begins with the coordinates of observations.
The observation points can be aggregated to regions within boundaries.
The polygon regions may be presented directly, or transformed first according to associated values.

The visualisation method used will depend on the intended use of the map, namely the message a map will communicate.

commonly used for hypothesis generating and as the first step in exploratory
spatial data analysis. @SE Geography is concerned with the identification and
explanation of spatial structure, pattern, and process, and with the analysis and explanation of the links
between humans and the environment (2). @SAMGIS

By providing a visual representation of cancer outcomes, geographic patterns of disease are able
to be identified and effectively addressed
-->

@@

<!--Choropleth map (Steff)-->

<!-- what is a choropleth map -->
A choropleth map is often used to display the spatial characteritics of a relationship, or variability, of measurements. 
As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO].
A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI].
Early versions of choropleth maps used symbols or patterns instead of colour.
@CPISACA discusses their use in visualising cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.
Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.
A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form. 
Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

@@

@@

In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
Choropleth maps can be useful devices for communicating information to public on a familiar map base.
A cancer atlas is a choropleth map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. @MACM provide the definition and reports that cancer atlases began with Haviland's maps in 1875.
The data collection methods of cancer mortality rates across regions, and the administrative control within regions lends itself to choropleth visualisation. 
The increasing development and use of disease maps can be attributed to the availability of geographic information system software [@SE].
arly work in US cancer atlases can be attributed to Burbank ???, and UK cancer atlases to Howe 1963 ???.
The choropleth maps presented levels via hatchings or dots on a black and white scale.
These atlases were key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work related exposures, and high risk diets @MACM.

@@

@@

@@

<!--
Create a table:
cofactors: socio demographic status

@@

<!-- Going from choropleth to cartogram 
the background variable assumed to be uniform (such as population density) can be adjusted so that it behaves as if uniformity were the case
@GOINO

A cartogram is an alternative map visual. 
An intention distortion is applied to a geographic map base,

A cartogram is a map purposely distorted so that its spatial properties represent quantities
not directly associated with position on the globe. As thematic maps, cartograms emphasize
the distribution of a variable by changing the area (or lengths) of objects on the map.
Our approach maintains continuity, and preserves
many local features of cartographic lines that provide visual clues to the identity of the distorted
objects.
@ACCAC

@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed map users as map creators. These systems are utilised depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)’.

-->
<!--
criticisms:
Certain limitations to the standard graphic techniques have also become apparent, however.
They are slow, and do not permit instant retrieval of information. They are untrustworthy when interpretations of spatial association are needed; there are both severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually.
-->

<!--
Urbanisation has pushed people closer together,
-in workplaces and in homes. 
-this issue is not with the size of the administrative areas,
- the increasing urbanisation is contributing to the changing density of population

A particular spatial structure includes
the individuals affected and how they are connected in
communities, as well as the dynamics of these communities
and their organization into larger units. @SAMGIS
-->

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however, if the land mass presented covers enough of the globe, there must be a transformation or distortion to display in 2D. The amount of distortion is related to the distance displayed @GAMP.
The world projections reflect the frequent distortions seen from altering perspectives. Choropleth maps will always be distorted if they cover enough of the globe, as will photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favourable distortion of the geography for presenting the set of spatial information.
Diagrams that do not specify a projection can be considered to have some unknown projection. <!-- ?? cite tobler.-->

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Waldo Tobler [-@TVSSS] explores many graphical techniques, and suggests there are particular methods for particular purposes. Cartograms provide an alternative visualisation method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
Cartograms may be seen as an extension of map transformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].

@TAAM variations in population density may obscure geographic patterns
This factor may be eliminated by making the map area proportional to population rather than to actual land area.
Such maps, or what have been termed "population-by-area cartograms," are well known in the
cartographic literature
-->
Using choropleth maps for population characteristics requires graphic distortions when the population concerned varies greatly in density [@CTTMB].
When implementing a distortion of the geographical shape according to population, an area cartogram [@NAC], or iso-demographic map is the result.
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just, giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. @CTTMB suggest that spatial socio-economic data is best presented on a cartogram for urban areas.
@HEDP agrees that 'cancer occurs in people, not in geographical areas' and the map bases of population reflect this and avoid allocating 'undue prominence' to rural areas. @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. Using health estimates from large areas can prevent drawing attention to the inequalities.

<!--
> The aim here is to see the whole, in as much detail as possible, at a glance.

@HEDP discusses the impact of electronic computer-assisted techniques. Disease atlases were implemented using both geographic (Howe 1963) and demographic bases (Howe 1970).

@AOCAFE
An Online Census Atlas for everyone
The familiar choropleth maps in the Online Census Atlas were produced using generalized boundaries. Great Britain is recognizable and it is relatively easy to identify the higher geography levels such as Counties and Local Authorities. A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.

The Atlas does not allow visualization of census
data at ward level, but instead uses these boundaries
aggregated to the six higher level geographies
defined above for display of 88 census variables.

To cartograms by:
The creation of the map began
with a computer-generated cartogram of parliamentary
constituencies where each constituency was
given equal weight.
Using the circular cartogram
algorithm (Dorling 1995), this resulted in a largely
hexagonal tessellation of 641 circles, each circle
representing a parliamentary constituency as first
used in the 1997 general election.
-->

@@

The area on the map space can also be used represent a value. There have been many algorithms presented, @SAIC provided a framework to investigate implementations and the "statistical accuracy, geographical accuracy, and topological accuracy". The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].

Cancer statistics and atlases

@@

<!-- Creation goals-->

'Firstly, that no such transformation has a unique solution, leaving the subjective choice in the hands of the individual map designer. '

1) that the size of the electoral subdivisions should be made proportional to their respective electoral populations, 
2) that the contiguity relationships between the DCU's should be maintained, 
3) that true directions should be retained from the centre of the transformation, 
4) that the main shape characteristics of the DCU's should be preserved, and 
5) that the map should possess the overall shape of the built-up area.

@@

<!-- Intentionally preserve shape-->

A contiguous cartogram allows the map space to highlight the distribution of the variable. These cartograms maintain connectivity of the map regions while areas are resized according to a statistic, this often must occur at the expense of the shape [@CBATCC,@NAC, @TAAM].
From a computer graphics perspective, @ACA believe it is a problem of 'map deformation' to account for the value assigned to each area, they provide three methods for creating value-by-area cartograms.
Examples include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].

@RSCW
helps to correct misconceptions. Pupils are likely to over-estimate regions as Canada, Siberia, Africa, and our own West just because they loom up large on maps. The simple picture which the cartogram gives is easily comprehended and remembered and
will be stimulating in making comparisons.

@NISCC
two advantages of the cartogram: (1) new relationships are often revealed visually, and ( 2 ) it is often more honest to use “other” space on the base map than land area. Finally, and most important for the present discussion, they suggest that the ability to preserve shape of the statistical units drops with an increase in the number of these units, and in situations where the difference is great between the smallest and largest values mapped.
Until we know how the map reader compares it to an ordinary thematic map, we are experimenting with a graphic device that may be reducing (or at least hindering) the flow of information from the map to the map reader.
-->

@@

constant shapes are easily recognized regardless of scale;
-->

@@

@ACTUC suggests 'population distribution is often extremely uneven in former British colonies'.

@@

To choose an approporiate map display, the map creator must consider the intended user, and message the map will communicate.
It is the objectives of the investigator that will drive the choice of representation [@CPISACA].

>>>>

Maps have been adopted to present geospatial statistics for centuries.
Maps connect data and statistics to the geographical representation of areas that are generally familiar to the audience.
However, it is not enough for areas on maps to be recognisable, if information regarding the distribution of the statistic cannot be gleaned.
In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer statistics.
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. 
This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes, and health outcomes more broadly.
Cancer statistics directly represent and relate to the people living within individual geographical areas.

The visualisation methods used to present cancer statistics will depend on the intended message and users of the map. 
Transforming statistics to visualisations considers individual observations or observation points aggregated into communities and geographical units.
Visualisations may be used for hypothesis generating and is often the first step in exploratory spatial data analysis. 
Disease maps help to present geographic patterns that may have been overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed.

@@

@@

<!-- what is a choropleth map -->
A choropleth map is used to display the spatial characteritics of a relationship, or variability, of measurements. 
As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO].
A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI].
Early versions of choropleth maps used symbols or patterns instead of colour.
@CPISACA discusses their use in visualising cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.
Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.
A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form. 
Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

@@

<!--
They are untrustworthy when interpretations of spatial association are needed; there are both severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually.
-->

@@

In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
Choropleth maps can be useful devices for communicating information to public on a familiar map base.
A cancer atlas is a choropleth map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. @MACM provide the definition and reports that cancer atlases began with Haviland's maps in 1875.
The data collection methods of cancer mortality rates across regions, and the administrative control within regions lends itself to choropleth visualisation. 
The increasing development and use of disease maps can be attributed to the availability of geographic information system software [@SE].
Early work in US cancer atlases can be attributed to Burbank ???, and UK cancer atlases to Howe 1963 ???.
The choropleth maps presented levels via hatchings or dots on a black and white scale.
These atlases were key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work related exposures, and high risk diets @MACM.

@@

An Online Census Atlas for everyone [@AOCAFE] uses familiar choropleth, produced using generalized boundaries for six higher level geographies. Great Britain is recognizable and it is relatively easy to identify the higher geography levels such as Counties and Local Authorities.

@@

<!--
'Where control of the message is important, static maps will continue to be the most effective, although good tables, graphs, and explanatory text are still needed in order to ensure that different people will see the same thing in the maps' @CPISACA

@@

<!--
Create a table:
cofactors: Greater capital city regions, states, socio demographic status, Remoteness

@@

> The aim here is to see the whole, in as much detail as possible, at a glance.

<!-- What is a cartogram? -->
Cartograms provide an alternative visualisation method for statistical and geographical information.
A cartogram alters the map base with the intention of improving the presentation of a statistic.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
For a single variable of interest, each map area is changed to emphasise the distribution by representing the corresponding value, in comparison to the value of the other areas [@ACCAC].

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D.
The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just as photographs of the globe from space.
Choropleth creation requires choosing a map projection that shows a favourable distortion of the geography for presenting the set of spatial information.
Diagrams that do not specify a projection can be considered to have some unknown projection. <!-- ??? cite tobler.-->

@@

<!-- Denominators -->
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed beyond choosing a projection. 
The purposeful distortion of the map space is beneficial when a uniform density of the map base is desired, @ACTUC suggests 'population distribution is often extremely uneven in former British colonies', this makes the distortion necessary. 
When a map base is transformed according to population density, population becomes a uniformly distributed background for the statistic presented [@GOINO]. 
Using choropleth maps for population characteristics requires graphic distortions when the population concerned varies greatly in density [@CTTMB].
When implementing a distortion of the geographical shape according to population, an area cartogram [@NAC], population-by-area cartograms [@TAAM], or iso-demographic map is the result.

A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult [@AOCAFE].
The distortion of a cartogram prevents the density obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just, giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@CTTMB suggest that spatial socio-economic data is best presented on a cartogram for urban areas.
@HEDP agrees that 'cancer occurs in people, not in geographical areas' and the map bases of population reflect this and avoid allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

<!--
@HEDP discusses the impact of electronic computer-assisted techniques. Disease atlases were implemented using both geographic (Howe 1963) and demographic bases (Howe 1970).

2) that the contiguity relationships between the DCU's should be maintained, 
3) that true directions should be retained from the centre of the transformation,
4) that the main shape characteristics of the DCU's should be preserved, and 
5) that the map should possess the overall shape of the built-up area. 
-->

@@

There have been many algorithms presented, @SAIC provided a framework to investigate implementations and the "statistical accuracy, geographical accuracy, and topological accuracy".

@@

@@

<!-- Intentionally preserve neighbours -->

A contiguous cartogram maintains connectivity of the map regions while areas are resized according to a statistic. This transformation often occurs at the expense of the shape [@CBATCC, @NAC, @TAAM].
From a computer graphics perspective, @ACA believe it is a problem of 'map deformation' to account for the value assigned to each area, they provide three methods for creating value-by-area cartograms.
Examples include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].

@NISCC
two advantages of the cartogram: (1) new relationships are often revealed visually, and ( 2 ) it is often more honest to use “other” space on the base map than land area. Finally, and most important for the present discussion, they suggest that the ability to preserve shape of the statistical units drops with an increase in the number of these units, and in situations where the difference is great between the smallest and largest values mapped.
Until we know how the map reader compares it to an ordinary thematic map, we are experimenting with a graphic device that may be reducing (or at least hindering) the flow of information from the map to the map reader.

The transformations required to create cartograms was historically the role of professional cartographers.  
Geographical information systems allowed map users, and researchers to implement their own cartograms, but these systems are utilised depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)’[@CD].

@@

constant shapes are easily recognized regardless of scale;

Presenting an alternative view can help to correct interpretation of global statistics when some areas loom large on maps [@RSCW].
-->

@@

@@

Waldo Tobler [-@TVSSS] explores many graphical techniques, and suggests there are particular methods for particular purposes. To choose an approporiate map display, the map creator must consider the intended user, and message the map will communicate.
It is the objectives of the investigator that will drive the choice of representation [@CPISACA].
